## Course Project for Reproducible Research
Athena Marquez | 14 March 2021

### Loading packages for this project
```{r loadpackages, echo=TRUE}
library(dplyr)
library(ggplot2)
```

### Loading and preprocessing the data
**The following code loads the data and transforms it into a format suitable for analysis.**
```{r processdata, echo=TRUE}
#Download and unzip file to working directory
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
download.file(url, './repdata_data_activity.zip')
unzip('./repdata_data_activity.zip')

#Read and view files
data <- read.csv('activity.csv')
```

### Total Number of Steps Taken Per Day

**1. This syntax shows the calculation of the total number of steps taken per day.**
```{r totalsteps, echo=TRUE}
day <- group_by(data, date) 
dailytotal <- summarize(day, steps = sum(steps, na.rm=TRUE))
print(dailytotal)
```


**2. This creates a histogram of the total number of steps taken per day.**
```{r histsteps, echo=TRUE}
hist(dailytotal$steps, col = "wheat", main = "Total Number of Steps Taken Per Day", xlab = "Total Steps Per Day")
```


**3. The following calculates and reports the mean and median of the total number of steps taken per day.**
```{r calculatedata, echo=TRUE}
dailystat <- summarize(day, mean = mean(steps, na.rm=TRUE), median = median(steps, na.rm=TRUE))
print(dailystat)
```

### Average Daily Pattern

**1. The code below creates a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.**
```{r timeseriesplot, echo=TRUE}
interval <- group_by(data, interval)
average <- summarize(interval, steps = mean(steps, na.rm=TRUE))
plot(average$interval, average$steps, xlab = "Interval", ylab = "Average Number of Steps Taken", main = "Average Daily Pattern", type = "l")
```

**2. The maximum average number of steps across all days occurs on which 5-minute interval?**
```{r whichinterval, echo=TRUE}
maxsteps <- average[which.max(average$steps),]
print(maxsteps)
```


### Imputing Missing Values

**1. Shown below is the code that calculates and reports the total number of missing values in the data set.**
```{r countNAs, echo=TRUE}
nrow(data[data$steps == "NA",])
```

**2. This code fills in all of the missing values in the dataset with the mean for each corresponding 5-minute interval.**
```{r filldata, echo=TRUE}
newdata <- data
for(i in 1:nrow(newdata)){
  if(is.na(data$steps[i])){
    newdata$steps[i] <- average$steps[match(data$interval[i], average$interval)]
  }
}
```

**3. This line shows the new dataset with the missing data filled in.**
```{r newdataset, echo=TRUE}
head(newdata)
```

**4. The following shows the histogram of the new data and reports the mean and median of the total number of steps taken per day.**
```{r newtotalsteps, echo=TRUE}
newday <- group_by(newdata, date) 
newdailytotal <- summarize(newday, steps = sum(steps, na.rm=TRUE))
print(newdailytotal)
hist(newdailytotal$steps, col = "wheat", main = "Total Number of Steps Taken Per Day", xlab = "Total Steps Per Day")
newdailystat <- summarize(newday, mean = mean(steps, na.rm=TRUE), median = median(steps, na.rm=TRUE))
print(newdailystat)
```
It can be seen that the original histogram differs from the new histogram in terms of shape, where the latter resembles a more bell-shaped curve. Imputing missing data has increased the estimates of the total daily number of steps as well.

### Differences in Activity Patterns Between Weekdays and Weekends

**1. This part creates a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given data is on a weekday or weekend day.**
```{r newvariable, echo=TRUE}
newdata$day <- ifelse(weekdays(as.Date(data$date))%in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "weekday", "weekend")
head(newdata)
```


**2. The code below shows the creation of a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.**
```{r newtimeseriesplot, echo=TRUE}
newaverage <- aggregate(steps~interval+day, newdata, mean) 
ggplot(newaverage, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) + labs(x ="Interval") + labs(y = "Average Number of Steps") + labs(title = "Average Daily Pattern - Weekdays vs. Weekends")

```




